The Emotional Coherence Controversy: Evidence from Human and Automatic Expression Recognition

### 2019-04-23 ACII_2019_paper/ACII_2019_paper.Rmd
submission

  institution:
    - name: Dublin City University
      department: Business School
      location: Dublin, Ireland
      email: damien.dupre@dcu.ie
      mark: 1
      author:
        - name: Damien Dupré
    - name: Université Grenoble Alpes
      department: Laboratoire Inter-universitaire de Psychologie
      location: Grenoble, France
      mark: 2
      author:
        - name: Anna Tcherkassof
          email: anna.tcherkassof@univ-grenoble-alpes.fr
abstract: |
  While it has been taken for granted in the development of several automatic facial expression recognition tools, the question of the coherence between subjective feelings and facial expressions is still a subject of debate. On one hand, the behaviorist approach conceives emotions as genetically hardwired and therefore being genuinely displayed through facial expressions. On the other hand, the constructivist approach conceives emotions as socially constructed; the emotional meaning of a facial expression being inferred by the observer. In order to evaluate the coherence between the subjective feeling of emotions and their recognition based on facial expression, 232 videos of encoders recruited to carry out an emotion elicitation task were annotated by 1383 human observers as well as by an automatic facial expression classifier. Results show low accuracy of human observers and of the automatic classifier to infer the subjective feeling from the facial expressions displayed by encoders. They also show a weak coherence between self-reported emotional states and facial emotional displays. Based on these results, the hypothesis of genetically hardwired emotion genuinely displayed is difficult to support, whereas the idea of emotion and facial expression socially constructed appears to be more likely. Accordingly, automatic emotion recognition tools based on facial expressions should be questioned.

@@

bibliography: mybibfile.bib
output: rticles::ieee_article
csl: http://www.zotero.org/styles/ieee
fig_caption: yes
keep_tex: yes
number_sections: yes
includes:
  in_header: preamble-latex.tex

template: C:/Users/dupred/Desktop/Projects/dynemo_feel_express/ACII_2019_paper/conference_041818.tex
---

\begin{IEEEkeywords}
Facial expression, self-report, human observer, automatic recognition.
\end{IEEEkeywords}

@@

## A. The Behaviorist Approach
Based on the behaviorist approach initiated by Darwin in *The Expression of the Emotions in Man and Animals* [@darwin1872expression], facial expressions are conceived as a genuine displays of an individual's inner emotional state. This hypothesis is used as a basis for the Basic Emotion Theory which states that a set of six emotions are universally displayed and are genetically hardwired not only in humans [@ekman1992argument] but also in different animal species [@de2019mama]. According to the view of the Basic Emotion Theory, *“when emotions are aroused by perception of a social event, a set of central commands produce patterned emotion-specific changes in multiple systems, including […] facial expressions.”* [@ekman2007directed, 49]. To respond to criticisms, several amendments have been made to the Basic Emotion Theory, increasing the number of basic emotions from six to seven [@ekman1988universality] as well as adding the concept of “display rules” to explain cultural differences in the management of facial expressions [@ekman1987universals].

@@

The authors would like to thank Brigitte Meillon and Jean Michel Adam who developed the software used to collect and preprocess human observer results.

>>>>

  institution:
    - name: #Dublin City University
      department: #Business School
      location: Blank for review
      email: #damien.dupre@dcu.ie
      mark: 1
      author:
        - name: #Damien Dupré
    - name: #Université Grenoble Alpes
      department: #Laboratoire Inter-universitaire de Psychologie
      location: Blank for review
      email: #anna.tcherkassof@univ-grenoble-alpes.fr
      mark: 2
      author:
        - name: #Anna Tcherkassof

abstract: |
  While it has been taken for granted in the development of several automatic facial expression recognition tools, the question of the coherence between subjective feelings and facial expressions is still a subject of debate. On one hand, the behaviorist approach conceives emotions as genetically hardwired and therefore being genuinely displayed through facial expressions. On the other hand, the constructivist approach conceives emotions as socially constructed; the emotional meaning of a facial expression being inferred by the observer. In order to evaluate the coherence between the subjective feeling of emotions and their recognition based on facial expression, 232 videos of encoders recruited to carry out an emotion elicitation task were annotated by 1383 human observers as well as by an automatic facial expression classifier. Results show low accuracy of human observers and of the automatic classifier to infer the subjective feeling from the facial expressions displayed by encoders. They also show a weak coherence between self-reported emotional states and facial emotional displays. Based on these results, the hypothesis of genetically hardwired emotion genuinely displayed is difficult to support, whereas the idea of emotion and facial expression socially constructed appears to be more likely. Accordingly, automatic emotion recognition tools based on facial expressions should be questioned.

@@

bibliography: mybibfile.bib
output: rticles::ieee_article
csl: http://www.zotero.org/styles/ieee
fig_caption: yes
number_sections: yes
includes:
  in_header: preamble-latex.tex
---
<!-- template: C:/Users/dupred/Desktop/Projects/dynemo_feel_express/ACII_2019_paper/conference_041818.tex -->
\begin{IEEEkeywords}
Facial expression, self-report, human observer, automatic recognition.
\end{IEEEkeywords}

@@

## A. The Behaviorist Approach
Based on the behaviorist approach initiated by Darwin in *The Expression of the Emotions in Man and Animals* [@darwin1872expression], facial expressions are conceived as a genuine displays of an individual's inner emotional state. This hypothesis is used as a basis for the Basic Emotion Theory which states that a set of six emotions are universally displayed and are genetically hardwired not only in humans [@ekman1992argument] but also in different animal species [@de2019mama]. According to the view of the Basic Emotion Theory, *“when emotions are aroused by perception of a social event, a set of central commands produce patterned emotion-specific changes in multiple systems, including […] facial expressions.”* [@ekman2007directed, 49]. To respond to criticisms, several amendments have been made to the Basic Emotion Theory, increasing the number of basic emotions from six to seven [@ekman1988universality] as well as adding the concept of “display rules” to explain cultural differences in the management of facial \linebreak expressions [@ekman1987universals].

@@

[blank for review]
<!-- The authors would like to thank Brigitte Meillon and Jean Michel Adam who developed the software used to collect and preprocess human observer results. -->

### 2019-04-23 ACII_2019_paper/ACII_2019_paper.tex
submission

%% ---- classic IEEETrans one column per institution --------
 %% -- beg if/affiliation.institution-columnar
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Damien Dupré %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Business School\\
Dublin City University\\
Dublin, Ireland
\\damien.dupre@dcu.ie
}
\and
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Anna Tcherkassof %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Laboratoire Inter-universitaire de Psychologie\\
Université Grenoble Alpes\\
Grenoble, France
  %% -- beg for/affiliation.institution.author
\\anna.tcherkassof@univ-grenoble-alpes.fr
 %% -- end for/affiliation.institution.author
}
 %% -- end for/affiliation.institution
 %% -- end if/affiliation.institution-columnar
%% ----------------------------------------------------------

@@

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} {[}2{]}, facial
expressions are conceived as a genuine displays of an individual's inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory which states that a set of six emotions are universally
displayed and are genetically hardwired not only in humans {[}3{]} but
also in different animal species {[}4{]}. According to the view of the
Basic Emotion Theory, \emph{``when emotions are aroused by perception of
a social event, a set of central commands produce patterned
emotion-specific changes in multiple systems, including {[}\ldots{}{]}
facial expressions.''} {[}5, p. 49{]}. To respond to criticisms, several
amendments have been made to the Basic Emotion Theory, increasing the
number of basic emotions from six to seven {[}6{]} as well as adding the
concept of ``display rules'' to explain cultural differences in the
management of facial expressions {[}7{]}.

@@

The authors would like to thank Brigitte Meillon and Jean Michel Adam
who developed the software used to collect and preprocess human observer
results.

>>>>

%% ---- classic IEEETrans one column per institution --------
 %% -- beg if/affiliation.institution-columnar
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
 %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{\\
Blank for review
  %% -- beg for/affiliation.institution.author
 %% -- end for/affiliation.institution.author
}
\and
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
 %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{\\
Blank for review
  %% -- beg for/affiliation.institution.author
 %% -- end for/affiliation.institution.author
}
 %% -- end for/affiliation.institution
 %% -- end if/affiliation.institution-columnar
%% ----------------------------------------------------------

@@

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} {[}2{]}, facial
expressions are conceived as a genuine displays of an individual's inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory which states that a set of six emotions are universally
displayed and are genetically hardwired not only in humans {[}3{]} but
also in different animal species {[}4{]}. According to the view of the
Basic Emotion Theory, \emph{``when emotions are aroused by perception of
a social event, a set of central commands produce patterned
emotion-specific changes in multiple systems, including {[}\ldots{}{]}
facial expressions.''} {[}5, p. 49{]}. To respond to criticisms, several
amendments have been made to the Basic Emotion Theory, increasing the
number of basic emotions from six to seven {[}6{]} as well as adding the
concept of ``display rules'' to explain cultural differences in the
management of facial \linebreak expressions {[}7{]}.

@@

{[}blank for review{]}
